---
title: "Demo: Visualization Techniques in `ggplot2`"
output:
  html_document:
    css: include/lab.css
    highlight: pygments
    theme: cerulean
    toc: true
    toc_float: true
---

```{r global_options, include = FALSE, warning = FALSE}
library(dplyr)
library(ggplot2)
```

In this walkthrough, we'll explore a bunch of standrd visualization types on a pair of simple datasets. For each visualization type, we'll start with a bare-bones example, and then move on to a more complex version to show off some of what `ggplot2` and its friends can do. Feel free to move at your own pace; play around with each vis technique on your own datasets, or if you're feeling adventurous take a crack at the **Challenge Problem** at the very end.

## Visualization Techniques

First we'll need some data. For this walkthrough, we'll use `mtcars` (which comes preinstalled with your `R` installation):
```{r}
head(mtcars)
```


### Bar Chart

-   Used for comparable variables
-   Compares quantitative values for different categories
-   Highlights relative amounts
-   Grouped/stacked bars can break each category into sub-groups

#### A simple Bar Chart in `ggplot2`
```{r}
mtcars %>%
  ggplot(aes(x = cyl)) + 
    geom_bar()
```

#### A slightly prettier Bar Chart with color and labels
```{r}
ggplot(mtcars, 
       aes(x = as.factor(cyl),             # Dual-encode x-value with color
           fill = as.factor(cyl))) +       #   (just because it's pretty)
  
  geom_bar() +                             # Add bars
  
  xlab("Number of cylinders") +            # Label the axes
  ylab("Count") +                          
  
  ggtitle("Count of cars by cylinders") +  # Add a title
  
  guides(fill = FALSE)                     # Remove the legend
```

### Histogram

-   Looks like a bar chart… but the x-axis is continuous
-   Y-axis shows count or relative frequency
-   Highlights distribution
-   Note: bin size makes a big difference!


#### A simple Histogram in `ggplot2`
```{r}
ggplot(mtcars, aes(x = wt)) + 
  geom_histogram()
```

#### A better Histogram with scaling, 10 bins, labels, and color
```{r}
ggplot(mtcars, aes(x = wt*1000)) +             # Scale wt to lbs
  
  geom_histogram(bins = 10,                    # Change the number of bins to 10
                 fill = rgb(0.1, 0.4, 0.5)) +  #    and fill with more aesthetically-pleasing color
  
  xlab("Weight (lb)") +                        # Label the axes
  ylab("Count") +                             
  
  ggtitle("Distribution of cars by weight")    # Add a title
```

### Boxplot

-   Also useful for highlighting distribution
-   Calls out key values: median, 1st & 3rd quartiles, “fences”, outliers
-   Use “jitter” to show actual values
-   Reference lines can help provide context
-   Can use annotations to show statistical significance

#### A simple boxplot in `ggplot2`
```{r}
ggplot(mtcars, aes(x = as.factor(cyl), y = mpg)) + 
  geom_boxplot()
```

#### A better boxplot with original points, semi-transparent fill, and labels
```{r}
library(ggpubr) # Useful library for making "publication-ready" graphics:
                # We'll use the stat_compare_means(...) function to add significance

ggplot(mtcars, aes(x = as.factor(cyl), y = mpg)) + 
  
  geom_boxplot(outlier.shape = NA,                    # Hide outliers so they don't double-plot
               fill=rgb(0.2, 0.4, 0.2, 0.6)) +        #    and change fill and opacity
  
  geom_jitter(width = 0.01,                           # Add original datapoints (jittered)
              alpha = 0.4) +                          #    using transparency to avoid overplotting
  
  geom_hline(yintercept = 30,                         # A reference line at y = 30
             linetype = "dashed") +                   #    and we'll make it dashed
  
  annotate("text", label = "mpg = 30",                # Always good to label reference lines,
           x = 3.3, y = 31)  +                        #    but boxplots have weird x-coords: 
                                                      #    1st category = 1, etc.
  
  xlab("Number of cylinders") +                       # Label the axes
  ylab("Fuel economy (mpg)") +                            
  
  ggtitle("Distribution of mpg by # Cylinders") +     # Add a title
  
  stat_compare_means(comparisons = list(c("4","6"),   # Add significance testing
                                         c("6","8")),
                      method = "t.test",
                      aes(label = ..p.signif..))
```

### Scatterplot

-   Shows the relationship between two continuous variables
-   Each point in the plot represents an observation
-   You can change color or symbol to highlight groups
-   Sometimes useful to show a trend line (regression)


#### A simple scatterplot in `ggplot2`
```{r}
ggplot(mtcars, aes(x = wt, y = mpg)) + 
  geom_point()
```

#### A better scatterplot with scaled values, shapes, color, and labels
```{r}
mtcars %>%
  mutate(wt = wt * 1000) %>%                         # Scale the x-values
  ggplot(aes(x = wt, y = mpg)) + 
  
    geom_point(size = 4,                             # Make the points a little bigger
               
             aes(shape = as.factor(cyl),             # Add an aesthetic to map #cyl to shape
                 color = as.factor(cyl))) +          #    dual encoded with color for visibility
  
  stat_smooth(method = "lm",                         # Add a quadratic regression line
              formula = y ~ x + I(x^2)) +
  
  labs(color = "# Cylinders",                        # Relabel color/shape so that the
       shape = "# Cylinders",                        #    legend looks nice
       x = "Weight (lb)",                            # Another way to label the axes
       y = "Fuel economy (mpg)") +                  
  
   ggtitle("Weight vs. fuel economy by # cylinders") # Add a title
```

### Scatterplot Matrix (SPLOM)

-   Scatterplots show the relationship between just two continuous variables at a time
-   We can combine multiple scatterplots into a matrix to show additional relationships



#### A simple SPLOM using `pairs()`
```{r}
pairs(~ mpg + wt + disp, data = mtcars)
```

#### A better SPLOM using `GGally`
```{r}
library(GGally) # Useful library that extends some of ggplot2's functionality
                # We'll use the ggpairs(...) function to make a classy SPLOM

ggpairs(mtcars, 
        columns = c("mpg","disp","hp"),         # Select specific columns to compare
        
        columnLabels = c("Fuel economy (mpg)",  # Add clearer names for each column
                         "Displacement",
                         "Horsepower"),
        
        aes(color = as.factor(cyl),            # Use color to denote # cylinders
            alpha = 0.4)) +                    # Set to semi-transparent
  
  ggtitle("Fuel economy vs. displacement vs. horsepower by # cylinders")  # Add a title
```

### Line Chart
-   Shows the trend in one variable, often over time
-   Multiple lines can show multiple variables, or the same variable for multiple observations (must have the same scale!)
-   Highlights “position switches”

Line charts don't really make sense for `mtcars`, so first we'll load a different dataset: Dr. Arbuthnot's baptism records from 1629-1710.
```{r}
library(dplyr)
arbuthnot <- read.csv("http://www.openintro.org/stat/data/arbuthnot.csv") %>%
  mutate(total = boys + girls)
```

#### A simple line chart using `ggplot`
```{r}
library(dplyr)

arbuthnot %>%
ggplot(aes(x = year, y = total)) +
  geom_line()
```

#### A better line chart, built with tidy data
```{r}
library(tidyr)

# First, let's tidy up our data so it's tall instead of wide
arbuthnot_tidy <- arbuthnot %>%
  select(-total) %>%
  pivot_longer(names_to = "gender", values_to = "count", -year)

ggplot(arbuthnot_tidy, aes(x = year, 
                           y = count, 
                           color = gender)) +     # Now we can map 'gender' to color to get 2 lines
  geom_line() +
  
  labs(color = "Gender",                          # Relabel color
       x = "Year",                                # Label the axes
       y = "Number of records") +                 
  ggtitle("Dr. Arbuthnot's Baptism Records from 1629-1710 (by Gender)") # Add a title
```

## Challenge Problem

Try building a [**parallel coordinates plot**](https://en.wikipedia.org/wiki/Parallel_coordinates) using `ggplot2` on a dataset of your choice! __Hint: you can accomplish this using some of the packages we've already loaded, or try building it on your own using `geom_line`s and some clever data wrangling!__

```{r}
# Your code here
```