---
title: "ggplot demo"
author: "R. Jordan Crouser"
date: "6/24/2019"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
```

## Playing with ggplot
```{r}
library(tidyr)
library(dplyr)
library(ggplot2)
glimpse(msleep)
```

## First attempt: basic scatterplot

How does `sleep_rem` compare to `sleep_total` across species?
```{r}
msleep %>%
  ggplot(aes(x = sleep_total, 
             y = sleep_rem)) +
  geom_point()
```

## Second attempt: color-coded scatterplot

Any relation to what they eat?
```{r}
msleep %>%
  ggplot(aes(x = sleep_total, 
             y = sleep_rem,
             color = vore)) +
  geom_point()
```

## Third attempt: who's who?

Who are those outliers?
```{r}
msleep %>%
  ggplot(aes(x = sleep_total, 
             y = sleep_rem, 
             color = vore)) +
  geom_text(aes(label=name), 
            nudge_y = 0.25, 
            size = 2)
```

## Final attempt: icons instead of color coding
Can I save space on the legend by using icons to encode `vore`?

```{r}
library(stringr) # for string manipulation
library(ggimage) # for custom icons
msleep %>%
  
  # Get rid of rows with missing `vore` values
  drop_na(vore) %>%
  
  # Slight re-encoding of `vore` values (adding "vore" to the end)
  #   and then adding an image column to map each species to the appropriate icon
  mutate(vore = paste(vore, "vore", sep = ""),
         image = paste(vore, ".png", sep = "")) %>%
  
  # And now we plot!
  ggplot(aes(x = sleep_total, 
             y = sleep_rem, 
             color = vore)) +
  
  # Add the image
  geom_image(aes(image = image), size=.03) +
  
  # And the text, nudged just a bit so it doesn't overlap the image
  geom_text(aes(label = name), 
            nudge_y = 0.25, 
            size = 2) +
  
  # Drop the legend
  theme(legend.position = "none")
```

## Bonus: `facet_grid` on `vore`
Can I fix the overplotting by faceting on `vore`?

```{r}
library(stringr) # for string manipulation
library(ggimage) # for custom icons
msleep %>%
  
  # Get rid of rows with missing `vore` values
  drop_na(vore) %>%
  
  # Slight re-encoding of `vore` values (adding "vore" to the end)
  #   and then adding an image column to map each species to the appropriate icon
  mutate(vore = paste(vore, "vore", sep = ""),
         image = paste(vore, ".png", sep = "")) %>%
  
  # And now we plot!
  ggplot(aes(x = sleep_total, 
             y = sleep_rem, 
             color = vore)) +
  
  # Add the image
  geom_image(aes(image = image), size=.03) +
  
  # And the text, nudged just a bit so it doesn't overlap the image
  geom_text(aes(label = name), 
            nudge_y = 0.25, 
            size = 2) +
  
  # Drop the legend
  theme(legend.position = "none") +
  
  facet_grid(~vore)
```