---
title: "Graphics with `ggplot2`"
output:
  html_document:
    css: include/lab.css
    highlight: pygments
    theme: cerulean
    toc: true
    toc_float: true
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
library(dplyr)
library(ggplot2)
```

## Introduction

**Goal**: by the end of this lab, you will be able to use `ggplot2` to build several different data-driven graphics.

### Setting up
**Remember**: before we can use a library like `ggplot2`, we have to load it:

```{r load-ggplot2-lib, message = FALSE, eval = TRUE}
library(ggplot2)
```


### Why `ggplot2`?

Advantages of `ggplot2`

-   consistent underlying **grammar of graphics** (Wilkinson, 2005)
-   plot specification at a high level of abstraction
-   very flexible
-   `theme` system for polishing plot appearance (more on this later)
-   mature and complete graphics system
-   many users, active mailing list

### What Is The Grammar Of Graphics?

**The big idea**: independently specify plot building blocks and combine them to create just about any kind of graphical display you want. Building blocks of a graph include:

-   data
-   aesthetic mappings
-   geometric objects
-   statistical transformations
-   scales
-   coordinate systems
-   position adjustments
-   faceting

Using `ggplot2`, we can specify different parts of the plot, and combine them together using the `+` operator.

### Example Data: `Housing prices`

Let's start by taking a look at some data on housing prices:

``` {r read-housing-data, eval = TRUE}
housing <- read.csv("http://www.science.smith.edu/~jcrouser/SDS192/landdata-states.csv", header = T)
glimpse(housing)
```

(Data from <https://www.lincolninst.edu/subcenters/land-values/land-prices-by-state.asp>)

## Geometric Objects and Aesthetics

### Geometric Objects (`geom`)

Geometric objects or `geoms` are the actual marks we put on a plot. Examples include:

-   points (`geom_point`, for scatter plots, dot plots, etc)
-   lines (`geom_line`, for time series, trend lines, etc)
-   boxplot (`geom_boxplot`, for, well, boxplots!)
-   ... and many more!

A plot must have at least one `geom`; there is no upper limit. You can add a `geom` to a plot using the `+` operator.

You can get a list of available geometric objects by looking at the package documentation, available through the help interface:

``` {r geom-help, eval=FALSE}
?ggplot2
```

or simply type `geom_<tab>` in any good R IDE (such as Rstudio) to see a list of functions starting with `geom_`.

### Aesthetic Mapping (`aes`)

In `ggplot2`-land, *aesthetic* means "something you can see". Examples include:

-   position (i.e., on the x and y axes)
-   color ("outside" color)
-   fill ("inside" color)
-   shape (of points)
-   line type
-   size

Each type of `geom` accepts only a subset of all aesthetics -- refer to the `geom` help pages to see what mappings each `geom` accepts. Aesthetic mappings are set with the `aes()` function, short for aesthetic. 


### Points

Now that we know about geometric objects and aesthetic mapping, we're ready to make our first `ggplot`: a scatterplot. We'll use `geom_point` to do this, which requires `aes` mappings for `x` and `y`; all others are optional. 


``` {r scatterplot-example, message = FALSE, eval = TRUE}
hp2013Q1 <- housing %>% filter(Date == 2013.25) 
ggplot(hp2013Q1,
       aes(y = Structure.Cost, x = Land.Value)) +
  geom_point()
```

### Lines

A plot constructed with `ggplot` can have more than one `geom`. In that case, the mappings established in the `ggplot()` call are plot defaults that can be added to or overridden. For example, we could add a regression line to our plot:

``` {r trend-line-example, message = FALSE, eval = TRUE}
p1 <- ggplot(hp2013Q1, aes(x = Land.Value, y = Structure.Cost))

p1 + geom_point(aes(color = Home.Value)) +
    geom_smooth(method="lm")
```

### Smoothers

Not all geometric objects are simple shapes -- the `smooth` `geom` includes both a line and a ribbon.

``` {r smoother-example, message = FALSE, eval = TRUE}
p1 +
  geom_point(aes(color = Home.Value)) +
  geom_smooth()
```

### Removing standard errors

``` {r smoother-noSE, message = FALSE, eval = TRUE}
p1 +
  geom_point(aes(color = Home.Value)) +
  geom_smooth(se=FALSE)
```

### Text

Each `geom` accepts a particualar set of mappings--for example `geom_text()` accepts a `labels` mapping.

``` {r text-sample, message = FALSE, eval = TRUE}
p1 + 
  geom_text(aes(label=State), size = 3)
```

### Aesthetic Mapping VS Assignment

Note that variables are mapped to aesthetics with the `aes()` function, while fixed aesthetics are set outside the `aes()` call. This sometimes leads to confusion, as in this example:

``` {r right-vs-wrong-aes, message = FALSE, eval = TRUE}
p1 +
  geom_point(aes(size = 2), # incorrect! 2 is not a variable
                color="red")   # this is fine -- turns all points red
```

### Mapping Variables to Other Aesthetics

Other aesthetics are mapped in the same way as `x` and `y` in the previous example.

``` {r other-aes-example, message = FALSE, eval = TRUE, warning=FALSE}
p1 +
  geom_point(aes(color = Home.Value, 
                 shape = region))
```


## Scales: Controlling Aesthetics

Aesthetic mapping (i.e., with `aes()`) only says that a variable should be mapped to an aesthetic. It doesn't say *how* that should happen. For example, when mapping a variable to *shape* with `aes(shape ` x)= you don't say *what* shapes should be used. Similarly, `aes(color ` z)= doesn't say *what* colors should be used. Describing what colors/shapes/sizes etc. to use is done by modifying the corresponding *scale*. In `ggplot2`, `scales` include:

-   `position`
-   `color` and `fill`
-   `size`
-   `shape`
-   `linetype`

Scales are modified with a series of functions using a `scale_<aesthetic>_<type>` naming scheme. Try typing `scale_<tab>` to see a list of scale modification functions.

### Common Scale Arguments

The following arguments are common to most `scales` in `ggplot2`:

-   `name`:  the first argument specifies the axis or legend title
-   `limits`: the minimum and maximum of the `scale`
-   `breaks`:  the points along the scale where labels should appear
-   `labels`: the text that appear at each break

Specific scale functions may have additional arguments; for example, the `scale_color_continuous` function has arguments `low` and `high` for setting the colors at the low and high end of the scale.

### Scale Modification Examples

Start by constructing a dotplot showing the distribution of home values by `Date` and `State`.

``` {r dotplot-example, message = FALSE, eval = TRUE}
p3 <- ggplot(housing,
             aes(x = State,
                 y = Home.Price.Index))

p4 <- p3 + geom_point(aes(color = Date),
                      alpha = 0.5,
                      size = 1.5,
                      position = position_jitter(width = 0.25, height = 0))
```

Now let's modify the `breaks` and `labels` for the `x` axis and color scales:

``` {r changing-scale-example, message = FALSE, eval = TRUE}
p4 + scale_x_discrete(name = "State Abbreviation") +
  scale_color_continuous(name = "",
                          breaks = c(1975.25, 1994.25, 2013.25),
                          labels = c(1971, 1994, 2013))
```

Next change the low and high values to `blue` and `red`:

``` {r changing-color-example, message = FALSE, eval = TRUE}
p4 +
  scale_x_discrete(name = "State Abbreviation") +
  scale_color_continuous(name = "",
                          breaks = c(1975.25, 1994.25, 2013.25),
                          labels = c(1971, 1994, 2013),
                          low = "blue", high = "red")
```

### Using different color scales

`ggplot2` has a wide variety of color `scales`; here is an example using `scale_color_gradient2` to interpolate between three different colors:

``` {r color-scale-midpoint-example, message = FALSE, eval = TRUE}
p4 +
  scale_color_gradient2(name = "",
                        breaks = c(1975.25, 1994.25, 2013.25),
                        labels = c(1971, 1994, 2013),
                        low = "blue",
                        high = "red",
                        mid = "gray60",
                        midpoint = 1994.25)
```

### Available Scales

Here's a (partial) combination matrix of available scales:

| **Scale**         | **Types**  | **Examples**                     |
|-------------------|------------|----------------------------------|
| `scale_color_`    | `identity`   | `scale_fill_continuous`   |
| `scale_fill_`     | `manual`     | `scale_color_discrete`    |
| `scale_size_`     | `continuous` | `scale_size_manual`       |
|                            | `discrete`   | `scale_size_discrete`     |
| `scale_shape_`    | `discrete`   | `scale_shape_discrete`    |
| `scale_linetype_` | `identity`   | `scale_shape_manual`      |
|                            | `manual`     | `scale_linetype_discrete` |
| `scale_x_`        | `continuous` | `scale_x_continuous`      |
| `scale_y_`        | `discrete`   | `scale_y_discrete`        |
|                            | `reverse`    | `scale_x_log`             |
|                            | `log`        | `scale_y_reverse`         |
|                            | `date`       | `scale_x_date`            |
|                            | `datetime`   | `scale_y_datetime`        |

**Note**: in RStudio, you can type `scale_` followed by TAB to get the whole list of available scales.

## Faceting

-   Faceting is `ggplot2` parlance for **small multiples**
-   The idea is to create separate graphs for subsets of data
-   `ggplot2` offers two functions for creating small multiples:
    1.  `facet_wrap()`: define subsets as the levels of a single grouping variable
    2.  `facet_grid()`: define subsets as the crossing of two grouping variables
-   Facilitates comparison among plots, not just of `geoms` within a plot

### Example: what is the trend in housing prices in each state?

Let's start by using a technique we already know -- map `State` to `color`:

``` {r line-plot-example, message = FALSE, eval = TRUE}
p5 <- ggplot(housing, aes(x = Date, 
                          y = Home.Value))

p5 + geom_line(aes(color = State))  
```

There are two problems here--there are too many `states` to distinguish each one by `color`, and the `lines` obscure one another.

### Faceting to the rescue!

We can fix the previous plot by `faceting` by `state` rather than mapping `state` to `color`:

``` {r faceting-example, message = FALSE, eval = TRUE}
(p5 <- p5 + geom_line() +
    facet_wrap(~State, ncol = 10))
```

There is also a `facet_grid()` function for faceting in two dimensions.

## The \#1 FAQ

### Map Aesthetic to Different Columns

The most frequently asked question goes something like this: *I have two variables in my data.frame, and I'd like to plot them as separate points, with different color depending on which variable it is. How do I do that?*

### First try: draw two separate lines

``` {r two-line-example, message = FALSE, eval = TRUE}
housing_byyear <- aggregate(cbind(Home.Value, Land.Value) ~ Date, data = housing, mean)

ggplot(housing_byyear,
        aes(x=Date)) +
  geom_line(aes(y = Home.Value), color="red") +
  geom_line(aes(y = Land.Value), color="blue")
```

This is not so great, because if I later add another dimension, I'll have to remember that I hard-coded these lines separately.

### A tidier way:

First we'll `gather` up the values we want to plot:
``` {r tidy-gather-example, message = FALSE, eval = TRUE}
library(tidyr)
home_land_byyear <- gather(housing_byyear,
                            value = "value",
                            key = "type",
                            Home.Value, Land.Value)
```

Then we'll plot them:
``` {r plot-gathered-data-example, message = FALSE, eval = TRUE}
ggplot(home_land_byyear,
        aes(x = Date,
            y = value,
            color = type)) +
  geom_line()
```

Much cleaner!

<div id="license">
This lab is based on the "[Introduction to R Graphics with `ggplot2`](http://tutorials.iq.harvard.edu/R/Rgraphics/Rgraphics.html)" workshop, which
is a product of the [Data Science Services](http://dss.iq.harvard.edu/) team [Harvard University](http://dss.iq.harvard.edu/). The original source is released under a 
[Creative Commons Attribution-ShareAlike 4.0 Unported](http://creativecommons.org/licenses/by-sa/4.0). 
This lab was adapted for SDS192: and Introduction to Data Science in Spring 2017 by R. Jordan Crouser at Smith College, and for the MassMutual data visualization workship in Summer 2017 by Amelia McNamara.
</div>